Type theory in type theory with single substitutions

### 2024-09-17 types_2024_postproc/p.tex
stuff
structure

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}).

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
level are included at all upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
type $A$ in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. The usual results from universal algebra (e.g.\ models
form a category with finite limits and initial object, existence of
free and co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreoever come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}). We define the notion
of model of a language (a SOGAT) by its parallel substitution
semantics, which is a GAT, and that induces a notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models, our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

The single substitution semantics of (\ref{eq:tytm}) is in some sense
too minimalistic: for example, consider a term $b$ which depends on a
context $\Gamma$ and an extra variable of type $A$, that is, $b :
\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
variable \emph{before} the last one obtaining $ b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
the end of the context, and $\p^+$ is weakening just before the last
variable in the context. Then, we instantiate the last $A[\p]$
variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary
model, but this equation holds in the syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we show that
the single substitution syntax of (\ref{eq:tt}) is equivalent to the
parallel substitution syntax of (\ref{eq:tt}), and the techniques that
we use here are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Equivalence of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tt}) than parallel substitution
models. The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax for monoid over $X$) are
the same as lists of $X$ (the syntax for nil/cons algebras over $X$).

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we always have to provide
a method for these operations, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions, all the
variables are instantiated. $\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In the first order single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT. Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from a single substitution model.
TODO: universe levels are a problem (maybe not if we use iterated
function space instead of $\Sigma$); the resulting CwF is not the same as
the free CwF over the SSC. Bootstrapping a CwF.
CwF -> SSC -> CwF = termification, that is equationally complete, see shallow paper, so we get an isomorphic CwF
SSC -> CwF -> SSC = we get a stronger thing than what we started with.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item Minimisation.
\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
  a parallel model from a single substitution model.
\end{itemize}

@@

TODO. In Section \ref{sec:tt}, we introduce the single substitution
calculus only relying on a working knowledge of an implementation of
type theory such as Agda, no prior experience with the metatheory of
type theory is required. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}), and we use this to
prove several admissible equations. Then we show how to derive the
rules of a CwF from our syntax (Section \ref{sec:cwf}). Finally, we
show that our syntax is isomorphic to the CwF-style syntax (Section
\ref{sec:iso}). We conclude in Section \ref{sec:conclusion}.

@@

Our metatheory in which this paper is
written is thus type theory, and everything is formalised in Agda.

@@

\section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}

@@

\begin{definition}[CwF]

>>>>

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the study of type systems
(e.g.\ \cite{DBLP:books/daglib/0005958}).

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
levels are included at upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
type $A$ in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2} and
our notation comes from two-level type theory \cite{DBLP:conf/csl/AltenkirchCK16,DBLP:journals/mscs/AnnenkovCKS23}). There is no good notion of homomorphism between
second-order models, thus the semantics of a SOGAT is given by a first
order generalised algebraic theory (GAT
\cite{DBLP:journals/apal/Cartmell86}). GATs are generalisations of
single-sorted algebraic theories (such as monoids, groups) to multiple
sorts where later sorts can be indexed over previous sorts. The usual
results from universal algebra (e.g.\ models form a category with
finite limits and initial object, existence of free and co-free
models) carry over to GATs \cite{andras,DBLP:phd/hal/Moeneclaey22}. An
example of a GAT is the theory of categories where there is a sort of
objects and a sort of morphisms double-indexed over objects.

@@

In the parallel semantics, the GAT always starts with a category with
a terminal object, objects of the category are called contexts,
morphisms are substitutions. Every sort/operation of the SOGAT gives
rise to a sort/operation in the GAT which is now indexed by
contexts. The sorts moreoever come with instantiation (sometimes
called substitution) operations which are functorial, and all the
operations are natural with respect to instantiation. For those sorts
which are bound by an operation in the SOGAT description, we have
context extension, meaning that contexts can carry variables of that
sort (in other words, the sort is given by a locally representable
presheaf). The parallel semantics of the SOGAT
\begin{equation}\label{eq:tytm}
\begin{alignedat}{10}
  & \Ty && : \Set \\
  & \Tm && : \Ty \ra\Set 
\end{alignedat}
\end{equation}
is the GAT of categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}, Definition \ref{def:cwf}).
We define the notion ``model of the SOGAT'' by its parallel substitution
semantics, which is a GAT, and that has an obvious notion of model. The
syntax of the language is the initial model, which exists for every
GAT \cite{DBLP:journals/pacmpl/KaposiKA19}. This is an intrinsic
syntax (only well-typed terms), quotiented by the equations in the
SOGAT, which describe conversion for the language.

The single substitution semantics is a GAT which does not start with a
category: there are contexts and maps between contexts, but these maps
are single substitutions or single weakenings, they don't
compose. There are fewer operations and fewer equations than in the
parallel semantics. In this paper we investigate the single
substitution semantics of the type theory specified by the SOGAT
(\ref{eq:tt}). This presentation of type theory is simpler than CwFs,
the operations are all easy to motivate, we illustrate this in Section
\ref{sec:tt}. Single substitution calculi are closer to the usual
presentation of type theory using derivation rules, this was one of
the motivations of Voevodsky for developing B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. Our single substitution
semantics of (\ref{eq:tytm}) is simpler and has more models than
B-systems. GAT presentations of type theory are important for defining
models. Our presentation gives a new notion of model, which in the
future might contribute to definitions of new models. It can also
influence implementations of type theory. Because our single
substitution GAT does not involve a category, it might be easier to
adapt it to a higher dimensional setting, see Section
\ref{sec:conclusion}.

The single substitution semantics is in some sense
too minimalistic: for example, consider a term $b$ which depends on a
context $\Gamma$ and an extra variable of type $A$, that is, $b :
\Tm\,(\Gamma\ext A)\,B$. We weaken $b$ so that we add an extra $A$
variable \emph{before} the last one obtaining $ b[\p^+] :
\Tm\,(\Gamma\ext A\ext A[\p])\,(B[\p]) $ where $\p$ is weakening at
the end of the context, and $\p^+$ is weakening just before the last
variable in the context. Then, we instantiate the last $A[\p]$
variable in the context by the previous $A$ obtaining
$b[\p^+][\langle\q\rangle] : \Tm\,(\Gamma\ext A)\,B$ where $\q$
denotes the last variable in the context (a.k.a\ De Bruijn index $0$).
We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an arbitrary single substitution
model, but this equation holds in the single substitution syntax.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold in the syntax. In fact, we show that
the single substitution syntax of (\ref{eq:tt}) is equivalent to the
parallel substitution syntax of (\ref{eq:tt}), and the techniques that
we use are not specific to the SOGAT (\ref{eq:tt}), we expect
that they work for any SOGAT. Equivalence of the syntaxes means that
the corresponding sorts ($\Con$, $\Ty$, $\Tm$) are isomorphic. For a
general SOGAT, there are more models of the correponding single
substitution GAT than the parallel substitution GAT. Every parallel
model is also a single model, but not necessarily the other way. While
B-systems are equivalent to CwFs (and C-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}), there are more single
substitution models of (\ref{eq:tytm}) than parallel substitution
models. The situation is similar to that of extensional combinatory
calculus and lambda calculus, where the syntaxes are equivalent, and
every lambda model is a combinatory model, but not the other way round
\cite{DBLP:conf/fscd/AltenkirchKSV23}. A simpler example is the
relationship of monoids over a set $X$ and nil/cons algebras over $X$,
where every monoid is a nil/cons-algebra, but not the other way; it is
well-known that free monoids over $X$ (syntax for monoid over $X$) are
the same as lists of $X$ (the syntax for nil/cons algebras over $X$).

When showing that the parallel and single substitution syntaxes are
equivalent, we define the parallel operations by induction on the
single substitution syntax. One of the operations in the single
substitution syntax is the instantiation operation: $\blank[\blank] :
\Ty\,\Gamma\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta$, and similarly for
terms. When doing induction on types/terms, we have to provide
a method for these operations, but this can be difficult as
substitutions don't compose. This is why we derive a new induction
principle which works on $\alpha$-normal types and terms. Just as
$\beta$-normal forms don't distinguish between $\beta$-equal terms,
$\alpha$-normal forms don't include explicit substitutions.
$\alpha$-normal forms are not usual normal
forms in the sense that they are still quotiented by the
computation/uniqueness rules like $\beta$/$\eta$ for function
space. To derive the new induction principle for the syntax, we first
prove $\alpha$-normalisation.

In the particular theory (\ref{eq:tt}), every type is represented by a
term. In its single substitution calculus, this can be
used to adapt some equations and remove some others, obtaining a more
minimalistic, but equivalent GAT.

Theory (\ref{eq:tt}) can be extended
with unit and $\Sigma$ types, more precisely with the following rules.
\begin{equation}\label{eq:sigma}
  \begin{alignedat}{10}
    & \top && : \Ty\,0 \\
    & \tt && : \Tm\,\top \\
    & \top\eta && : t = \tt \\
    & \Sigma && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
    & \fst,\snd && : \Tm\,(\Sigma\,A\,B)\cong(a:\Tm\,A)\times\,\Tm\,(B\,a): \blank,\blank
\end{alignedat}
\end{equation}
Using the same insight about types and the universe, the fact that
dependences can be represented by function spaces, and that
collections of dependencies can be collected using $\Sigma$ types, we
can actually derive a parallel model from any single substitution
model. This way we show that for the particular theory (\ref{eq:tt})+(\ref{eq:sigma}),
single and parallel substitution models are actually equivalent in a
weak sense.

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item For type theory with $\Pi$, $\U$, a minimised presentation of
  the equations which results in an isomorphic theory.
\item For type theory with $\Pi$, $\U$, $\top$, $\Sigma$, we show that
  single and parallel models are equivalent in a weak sense.
\end{itemize}

@@

In Section \ref{sec:tt}, we introduce the single substitution calculus
(SSC) only relying on a working knowledge of an implementation of type
theory such as Agda, no prior experience with the metatheory of type
theory is required. Section \ref{sec:admissible} shows that in the
syntax of SSC, several new equations are admissible, and that the
syntax of SSC is actually isomorphic to the CwF-based syntax. In
Section \ref{sec:minimisation} we minimise our SSC obtaining a theory
with fewer equations, relying on the presence of $\Pi$ and $\U$ in our
theory. In Section \ref{sec:cwf}, we show that in the presence of
$\Pi$, $\U$ and $\Sigma$, SSC-models are actually equivalent to
CwF-models in a weak sense. We conclude in Section
\ref{sec:conclusion}.

@@

Our metatheory in which this paper is
written is thus type theory, some sections are formalised in Agda.

@@

\section{CwF from SSC with $\Sigma$, $\Pi$ and $\U$}
\label{sec:cwf}

@@

\begin{definition}[CwF]\label{def:cwf}
